---
title: "Visualizing Microplastics"
subtitle: "EDS 240 Final Project Graphs" 
author: Isabella Segarra 
date: "3/13/26"
format: html
---

## Microplastics in the San Francisco Bay Estuary

This analysis utilizes data from San Francisco Estuary Institute (SEFI) regional study of microplastic pollution of a major urban estuary and adjacent ocean environment, sampling in stormwater, treated wastewater, surface water, sediment, and prey fish. In this document is the final versions of my visualizations for the EDS 240 Infographic.

## My questions

I am asking the following questions with my data:

+------------------------------------------------------------------+------------------------------------------------------------------------------+
| Overarching Question                                             | What are the sources and composition of microplastics in the SF Bay Estuary? |
+==================================================================+==============================================================================+
| **Question**                                                     | **Variables Used**                                                           |
+------------------------------------------------------------------+------------------------------------------------------------------------------+
| What types of plastic dominate in each source?                   | `sample_medium` × `plastic_type_refined`                                     |
+------------------------------------------------------------------+------------------------------------------------------------------------------+
| What is the size distribution of particles across plastic types? | `plastic_type` × `area (mm²)`                                                |
+------------------------------------------------------------------+------------------------------------------------------------------------------+
| What plastic types are found in fish tissue?                     | `` `sample_medium` `` × `plastic_type_refined`                               |
+------------------------------------------------------------------+------------------------------------------------------------------------------+
| What sampling environments had the most microplastics?           | `sample_medium`                                                              |
+------------------------------------------------------------------+------------------------------------------------------------------------------+
| What are the dominant colors of plastic?                         | `color`                                                                      |
+------------------------------------------------------------------+------------------------------------------------------------------------------+

```{r}
#| eval: false
#| echo: true
#| code-fold: true

##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##                                                                            --
##--------------------------- #......SET UP......-------------------------------
##                                                                            --
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

library(tidyverse)
library(tidyr)
library(here)
library(dplyr)
library(janitor)
library(readxl)
library(ggplot2)
library(stringr)
library(ggalluvial)
library(ggtext)
library(showtext)
library(glue)
library(packcircles)
library(ARTofR)
library(ggridges)
library(ggforce)

#.......Import data......

micro_plastics <- read_excel(here("data/2020-09-11_microparticledata.xlsx")) %>% 
  # Change all columns to lower_snack_case 
  clean_names() 

#.....Clean data with function....
source(here("clean_data_function.R"))

micro_plastics_clean <- clean_microplastics(micro_plastics)

#.......Fonts.......
font_add_google("Hanken Grotesk", "hanken")

#.....Establish palettes.....
my_pal <- c(
"Other"                                    = "#168aad",
"Common Synthetic Plastics"                = "#76c8c8",
"Industrial & Specialty Polymers"          = "#f4a261",
"Natural Materials"                        = "#99c1a9",
"Rubber"                                   = "#e9c46a",
"Synthetic Fibers & Textiles"              = "#9B2226",
"Unidentified plastic"                     = "#6c757d"
)

# This palette is based on the two groupings: common synthetic plastics and synthetic fibers and textiles 
plastic_pal <- c(
  "Polyethylene"  = "#76C8C8",  
  "Polypropylene" = "#9B2226",  
  "Polystyrene"   = "#76C8C8",  
  "Polyester"     = "#9B2226",  
  "Acrylic"       = "#9B2226"   
)
#........Enable showtest.....
showtext_auto(enable = TRUE)

```

## What plastic types dominate in each sampling environment?

```{r}
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##                                                                            ~~
##                               ALLUVIAL GRAPH                             ----
##                                                                            ~~
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

alluvial_plot <- micro_plastics_clean %>%
  filter(material_type_refined != "Unidentified plastic") %>% 
  filter(!is.na(material_type_refined), !is.na(sample_medium)) %>%
  count(sample_medium, material_type_refined) %>%
  # Plot 
  ggplot(aes(axis1 = material_type_refined, axis2 = sample_medium, y = n)) +
  geom_alluvium(aes(fill = material_type_refined),
                alpha = 0.8, 
                curve_type = "cubic") +
  scale_fill_manual(values = my_pal) +
  geom_stratum(aes(fill = material_type_refined), width = 1/3, color = "#002d3d") +
 geom_text(stat = "stratum", aes(label = after_stat(stratum)),
          family = "hanken", color = "#002d3d", size = 4) +
  scale_x_discrete(limits = c("Plastic type", "Sample medium"), expand = c(0.12, 0.1)) +
  labs(x = NULL, y = NULL, fill = "Plastic Type", 
       title = "Sources of Microplastics in the San Francisco Bay Estuary",
       subtitle = "Rigid synthetic polymers dominate microplastic composition across all sampling locations.", 
       caption = "Data: San Francisco Bay Estuary Institute") +
  theme_void(base_size = 20) +
  theme( 
        legend.position = "bottom",
        legend.title = element_text(color = "white", family = "hanken", size = 12),
        legend.text = element_text(color = "white", family = "hanken", size = 11),
        plot.title = element_text(color = "white", family = "hanken", 
                              face = "bold", 
                              hjust = 0.5, 
                              size = 18),
        plot.subtitle = element_text(color = "white", family = "hanken", hjust = 0.5, size = 13),
        plot.caption = element_text(color = "white", hjust = 1, family = "hanken"), 
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x = element_text(color = "white", family = "hanken", size = 4),
        panel.grid.major.y = element_blank(), 
        plot.margin = margin(t= 1, r = 1, b =1, l = 1, "cm"), 
        panel.border = element_blank(),
         panel.background = element_rect(fill = "#002d3d", color = NA),
    plot.background = element_rect(fill = "#002d3d", color = NA)
        ) 
# View plot
alluvial_plot

# Save plot as pdf
ggsave(here("plots", "alluvial.pdf"), alluvial_plot, 
       width = 10, height = 6, 
       device = cairo_pdf)

# Save plot as png 
ggsave(here("plots", "alluvial.png"), 
       plot = alluvial_plot, 
       width = 16, 
       height = 11,
       unit = "in",
       dpi = 300)

```

## What are the top plastics found in each sampling medium?

```{r}
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##                                                                            ~~
##                               BUBBLE CHART                            ----
##                                                                            ~~
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

# Find the top plastics 
top_plastics <- micro_plastics_clean %>%
  # Filter out "natura" materials 
  filter(!plastic_type %in% c("Not Characterized", "Unknown",
                               "Anthropogenic (unknown base)",
                               "Anthropogenic (cellulosic)",
                               "Unknown Potentially Rubber",
                               "Anthropogenic (synthetic)",
                               "Inorganic natural material",
                               "Stearates, Lubricants, Waxes",
                               "Cotton")) %>%
  count(plastic_type, sort = TRUE) %>%
  head(5) %>%
  mutate(circleProgressiveLayout(n, sizetype = "area"),
         label = paste0(plastic_type))

top_plastics_plot <- ggplot(top_plastics, aes(x0 = x, y0 = y)) +
  geom_circle(aes(r = radius, fill = plastic_type), color = NA) +
  geom_text(aes(x = x, y = y, label = label),
            color = "white", family = "hanken", size = 3.5) +
  coord_equal() +
  scale_fill_manual(values = plastic_pal) +
  theme_void() +
  labs(title = "Top plastics found in San Francisco Bay") + 
  theme(
    legend.position = "none",
    plot.title = element_text(color = "white", family = "hanken", hjust = 0.5),
    panel.grid.major = element_blank(),
    plot.background = element_rect(fill = "#002d3d", color = NA),
    panel.background = element_rect(fill = "#002d3d", color = NA)
  )

top_plastics_plot

# Save plot as pdf
ggsave(here("plots", "top_plastics_plot.pdf"), top_plastics_plot, 
       width = 10, height = 6, 
       device = cairo_pdf)

# save as png
ggsave(here("plots", "top_plastics_plot.png"), top_plastics_plot) 


```

## What percentage of each plastic end up in fish tissue?

```{r}
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##                                                                            ~~
##                               BAR CHART                            ----
##                                                                            ~~
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

fish_pct_plot <- micro_plastics_clean %>%
  filter(sample_medium == "Fish", !is.na(material_type_refined),
         material_type_refined != "Unidentified plastic") %>%
  count(material_type_refined) %>%
mutate(pct = n / sum(n) * 100,
       material_type_refined = fct_reorder(material_type_refined, pct)) %>% 
  ggplot(aes(x = 1, y = pct, fill = material_type_refined)) +
  geom_col() +
  geom_text(aes(label = paste0(round(pct, 0), "%")),
            position = position_stack(vjust = 0.5),
            color = "white", family = "hanken", size = 3.5) +
  scale_fill_manual(values = my_pal) +
  coord_flip() +
  labs(x = NULL, y = "Percentage (%)",
       title = "Types of plastic found in fish tissue",
       fill = "Plastic Type",
       caption = "Data Source: San Francisco Bay Estuary Institute (SFEI)") +
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    plot.title = element_text(color = "white", family = "hanken", face = "bold", hjust = 0.5),
    axis.text.x = element_text(color = "white", family = "hanken"),
    axis.title.x = element_text(color = "white", family = "hanken"),
    legend.text = element_text(color = "white", family = "hanken"),
    legend.title = element_text(color = "white", family = "hanken"), 
    panel.grid = element_blank(),
    plot.caption = element_text(color = "white", family = "hanken"),
    panel.background = element_rect(fill = "#002d3d", color = NA),
    plot.background = element_rect(fill = "#002d3d", color = NA)
  )
# View plot 
fish_pct_plot

# Save plot as pdf 
ggsave(here("plots", "fish_pct_plot.pdf"), fish_pct_plot, 
       width = 10, height = 6, 
       device = cairo_pdf)

# Save plot as png 
ggsave(here("plots", "fish_pct_plot.png"), fish_pct_plot)
```

## How much plastic was found in each sampling environment?

```{r}
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##                                                                            ~~
##                          TOP PLASTICS BAR CHART                        ----
##                                                                            ~~
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

total_n_plastics <- micro_plastics_clean %>%
  filter(!is.na(sample_medium)) %>%
  count(sample_medium, sort = TRUE) %>%
  mutate(
    sample_medium = fct_reorder(sample_medium, n)
  ) %>%
  ggplot(aes(x = sample_medium, y = n)) +
  geom_col(fill = "white") +
  geom_text(aes(label = scales::comma(n)), 
            hjust = -0.2, color = "white", 
            family = "hanken", size = 3.5) +
  labs(x = "Sampling environment", 
       y = "Number of microplastics found", 
       title = "What sampling environments had the most microplastics?",
       caption  = "Data Source: San Francisco Bay Estuary Institute (SFEI)") + 
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  coord_flip() +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "none",
    axis.title.y = element_text(color = "white", family = "hanken"),
    axis.text.y  = element_text(color = "white", family = "hanken"),
    plot.title = element_text(color = "white", family = "hanken", face = "bold", hjust = 0.5),
    axis.text = element_text(color = "white", family = "hanken"),
    axis.title.x = element_text(color = "white", family = "hanken"),
    plot.caption = element_text(color = "white", family = "hanken"),
    panel.grid = element_blank(),
    panel.background = element_rect(fill = "#002d3d", color = NA),
    plot.background = element_rect(fill = "#002d3d", color = NA)
  )

# View plot
total_n_plastics

# Save as pdf
ggsave(here("plots", "bar_chart.pdf"),total_n_plastics, 
       width = 10, height = 6, 
       device = cairo_pdf)

# Save as png

ggsave(here("plots", "bar_chart.png"), total_n_plastics)
```

## What is the size distribution of plastics by plastic type?

```{r}
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##                                                                            ~~
##                          DENISTY RIDGE PLOT                              ----
##                                                                            ~~
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

density_ridge_plot <- micro_plastics_clean %>%
  filter(!is.na(material_type_refined), !is.na(area_mm), area_mm > 0) %>%
  group_by(sample_medium) %>%
  filter(material_type_refined != "Unidentified plastic") %>% 
  filter(n() > 30, area_mm <= quantile(area_mm, 0.95)) %>%
  ungroup() %>%
  ggplot(aes(x = log10(area_mm), y = material_type_refined, fill = material_type_refined)) +
  geom_density_ridges2(alpha = 1, color = NA) +
  scale_fill_manual(values = my_pal) +
  scale_x_continuous(labels = function(x) scales::label_number()(10^x)) +  # convert log back to readable labels
  labs(
    title    = "What is the size distribution of plastics by plastic type",
    #subtitle = subtitle,
    x        = "Plastic particle area (mm²)",
    y        = NULL,
    caption  = "Data Source: San Francisco Bay Estuary Institute (SFEI)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.y  = element_text(color = "white", family = "hanken", size = ),
    axis.title.x = element_text(color = "white", family = "hanken",margin = margin(t = 15)),
    axis.text.x  = element_text(color = "white", family = "hanken", margin = margin(b = 5)),
    plot.caption = element_text(color = "white", family = "hanken"),
    plot.title   = element_text(color = "white", family = "hanken", face = "bold", hjust = 0.5),
    plot.subtitle = element_markdown(color = "white", family = "hanken", hjust = 0.5),
    legend.position = "none",
    panel.grid = element_blank(),
    panel.background = element_rect(fill = "#002d3d", color = NA),
    plot.background  = element_rect(fill = "#002d3d", color = NA)
    )
# View plot
density_ridge_plot

# Save as pdf
ggsave(here("plots", "bar_chart.pdf"),total_n_plastics, 
       width = 10, height = 6, 
       device = cairo_pdf)

# Save as png

ggsave(here("plots", "bar_chart.png"), total_n_plastics)

```

```{r}
# Turn off showtext
showtext_auto(enable = FALSE)
```
